Follow-up Samples

### 2017-09-15 submission/manuscript_R2.tex

\textbf{Background.} Despite growing evidence that members of the gut
microbiota can drive tumorigenesis, little is known about what happens
to it after treatment for an adenoma or carcinoma. This study tests
whether treatment alters the abundance of bacterial populations
associated with disease versus those with a normal colon by sequencing
the 16S rRNA genes in feces of 67 individuals before and after treatment
(adenoma n = 22), advanced adenoma n = 19), and carcinoma n = 26).

\textbf{Results.} The communities from patients with carcinomas changed
significantly more than those with adenoma following treatment (P-value
\textless{} 0.001). Although Random Forest models classified pre and
post-treatment samples for each diagnosis group, there was little
overlap between the OTUs that were indicative of each treatment.
Training Random Forest models on a larger cohort that contained
individuals with normal colons, adenomas, advanced adenomas, and
carcinomas we assessed the positive probability of having one of three
diagnoses and whether the pre and post-treatment samples were classified
the same. Patients who had carcinomas changed towards a microbial milieu
that resembles the normal colon after treatment (P-value \textless{}
0.05). There was no significant difference in the microbiota of
individuals treated with surgery alone and those treated with
chemotherapy or chemotherapy and radiation (P-value \textgreater{}
0.05).

\textbf{Conclusions.} It was difficult to identify significant
differences between pre and post-treatment samples from patients with
adenoma and advanced adenoma. However, for carcinomas, not only were
there large changes in pre versus post-treatment samples, but these
changes were towards a more normal microbiota.

\subsection{Importance}\label{importance}

This study helps to clarify what happens to the microbiota after
generalized treatment for an adenoma or carcinoma. It is a necessary
first step to ultimately investigating how the microbiota changes in
those with carcinoma recurrence. This is important since a large number
of those treated for carcinoma will have a recurrence. Finding better
ways to stratify those most and least likely to recur using the
microbiota could provide a cheaper and more accurate alternative to the
current status quo. Our results suggest that the microbiota could be
used to stratify individuals for risk of recurrence and provides a
starting point for future studies into this area.

@@

\textbf{\emph{The treatment of lesions are not consistent across
diagnosis groups.}} We used two approaches to identify those bacterial
populations that change between the two samples for each diagnosis
group. First, we sought to identify individual OTUs that could account
for the change in overall community structure. However, using a paired
Wilcoxon test we were unable to identify any OTUs that were
significantly different in the pre and post-treatment groups (P-value
\textgreater{} 0.05). It is likely that high inter-individual variation
and the irregular distribution of OTUs across individuals limited the
statistical power of the test. We attempted to overcome these problems
by using Random Forest models to identify collections of OTUs that would
allow us to differentiate between pre and post-treatment samples from
each of the diagnosis groups. The models overall performed moderately,
with the carcinoma model being the best (adenoma AUC range = 0.54 -
0.83, advanced adenoma AUC range = 0.34 - 0.65, carcinoma AUC range =
0.82 - 0.98). Interestingly, the top 10 most important OTUs by MDA that
were used for each model had little overlap with each other {[}Figure
2{]}. Although treatment had an impact on the overall community
structure, the effect of treatment was not consistent across patients
and diagnosis groups. Both the adenoma and carcioma treatment models had
AUCs that were significantly higher than a random model permutation
(P-value \textless{} 0.0001).

@@

\textbf{\emph{Model Building.}} The Random Forest (33) algorithm was
used to create the three models used to classify pre and post-treatment
samples by diagnosis (adenoma, advanced adenoma, or carcinoma). The
total number of individuals in the pre versus post-treatment models was
67 individuals. There were a total of 22 individuals in the pre versus
post-treatment adenoma model, 19 individuals in the pre versus
post-treatment advanced adenoma model, and 26 individuals in the pre
versus post-treatment carcinoma model {[}Table 1{]}.

Similarly, the Random Forest (33) algorithm was also used to create the
three models used to classify normal versus diagnosis. These samples
were obtained using the same methodology as described earlier in this
section. All samples used for this component of model training were from
pre-treatment samples only. The total number of individuals in the
normal versus diagnosis models was 423 individuals {[}Table 2{]}. There
were a total of 239 individuals in the normal versus adenoma model, 262
individuals in the normal versus advanced adenoma model, and 266
individuals in the normal versus carcinoma model {[}Table 2{]}.

All models included only OTU data obtained from 16S rRNA sequencing and
were processed and cleaned using the R package caret (v6.0.76).
Optimization of the mtry hyper-parameter involved making 100 different
80/20 (train/test) splits of the data where the same proportion was
present within both the whole data set and the 80/20 split. For each of
the different splits, 20 repeated 10-fold cross validation was performed
on the 80\% component to optimize the mtry hyper-parameter by maximizing
the AUC (Area Under the Curve of the Receiver Operator Characteristic).
The resulting model was then tested on the hold out data obtained from
the 20\% component. For adenoma pre versus post-treatment models the
optimized mtrys were 2 (n = 42), 42 (n = 26), and 914 (n = 32) for
adenoma. For advanced adenoma the mtrys were 2 (n = 26), 52 (n = 14),
and 1353 (n = 60). For carcinoma the mtrys were 2 (n = 36), 41 (n = 54),
and 863 (n = 10). For the normal versus diagnosis models the optimized
mtrys were (n = NA), (n = NA), and (n = NA) for adenoma. For advanced
adenoma (n = NA), (n = NA), and (n = NA). For carcinoma (n = NA), (n =
NA), and (n = NA). The hyper-parameter, mtry, defines the number of
variables to investigate at each split before a new division of the data
was created with the Random Forest model (33).

The treatment models were then assessed for their strength to classify
whether individuals had obtained treatment. We also assessed whether the
models performed better than random by creating 100 permutations of a
data set in which the labels were randomly assigned. To assess
similarity of the treatment models we looked for overlap between their
top 10 most important OTUs.

For the diagnosis models that were trained on a completely independent
sample set, each model was then applied to our 67-person cohort {[}Table
1{]} based on diagnosis: adenoma (pre-treatment adenoma (adenoma n = 22
and disease free n = 0) versus post-treatment adenoma (adenoma n = 0 and
disease free n = 22)), advanced adenoma pre-treatment advanced adenoma
(advanced adenoma n = 19 and disease free n = 0 ) versus post-treatment
advanced adenoma (advanced adenoma n = 0 and disease free n = 19), and
carcinoma (pre-treatment carcinoma (carcinoma n = 26 and disease free n
= 0) versus post-treatment carcinoma (carcinoma n = 1 and disease free n
= 25)). The application of the pre versus post-treatment models
generated the probabilities that the sample was a pre-treatment sample.
The application of the normal versus diagnosis models generated the
probabilities that the sample was that specific diagnosis (adenoma,
advanced adenoma, or carcinoma). We also looked at similarities between
the models by looking for overlap in the top 10\% of important OTUs to
each respective model.

>>>>

\textbf{Background.} Colorectal cancer is a worldwide health problem.
Despite growing evidence that members of the gut microbiota can drive
tumorigenesis, little is known about what happens to it after treatment
for an adenoma or carcinoma. This study tested the hypothesis that
treatment for adenoma or carcinoma alters the abundance of bacterial
populations associated with disease to those associated with a normal
colon. We tested this hypothesis by sequencing the 16S rRNA genes in the
feces of 67 individuals before and after treatment for adenoma (N = 22),
advanced adenoma (N = 19), and carcinoma (N = 26).

\textbf{Results.} There were small changes to the bacterial community
associated with adenoma or advanced adenoma and large changes associated
with carcinoma. The communities from patients with carcinomas changed
significantly more than those with adenoma following treatment (P-value
\textless{} 0.001). Although treatment was associated with intrapersonal
changes, the change in the abundance of individual OTUs in response to
treatment was not consistent within diagnosis groups (P-value
\textgreater{} 0.05). Because the distribution of OTUs across patients
and diagnosis groups was irregular, we used the Random Forest machine
learning algorithm to identify groups of OTUs that could be used to
classify pre and post-treatment samples for each of the diagnosis
groups. Although the three models could differentiate between the pre
and post-treatment samples, there was little overlap between the OTUs
that were indicative of each treatment. To determine whether individuals
who underwent treatment were more likely to have OTUs associated with
normal colons we used a larger cohort that contained individuals with
normal colons and those with adenomas, advanced adenomas, and
carcinomas. We again built Random Forest models and measured the change
in the positive probability of having one of the three diagnoses to
assess whether the post-treatment samples received the same
classification as the pre-treatment samples. Samples from patients who
had carcinomas changed towards a microbial milieu that resembles the
normal colon after treatment (P-value \textless{} 0.05). Finally, we
were unable to detect any significant differences in the microbiota of
individuals treated with surgery alone and those treated with
chemotherapy or chemotherapy and radiation (P-value \textgreater{}
0.05).

\textbf{Conclusions.} By better understanding the response of the
microbiota to treatment for adenomas and carcinomas, it is likely that
biomarkers will eventually be validated that can be used to quantify the
risk of recurrence and the likelihood of survival. Although it was
difficult to identify significant differences between pre and
post-treatment samples from patients with adenoma and advanced adenoma,
this was not the case for carcinomas. Not only were there large changes
in pre versus post-treatment samples for those with carcinoma, but these
changes were towards a more normal microbiota.

\subsubsection{Keywords}\label{keywords}

microbiota; colorectal cancer; polyps; treatment; risk factor.

@@

\textbf{\emph{The treatment of lesions are not consistent across
diagnosis groups.}} We used two approaches to identify those bacterial
populations that change between the two samples for each diagnosis
group. First, we sought to identify individual OTUs that could account
for the change in overall community structure. However, using a paired
Wilcoxon test we were unable to identify any OTUs that were
significantly different in the pre and post-treatment groups (P-value
\textgreater{} 0.05). It is likely that high inter-individual variation
and the irregular distribution of OTUs across individuals limited the
statistical power of the test. We attempted to overcome these problems
by using Random Forest models to identify collections of OTUs that would
allow us to differentiate between pre and post-treatment samples from
each of the diagnosis groups. The adenoma and carcinoma models performed
well (adenoma AUC range = 0.54 - 0.83 and carcinoma AUC range = 0.82 -
0.98); however, the model for patients treated for advanced adenomas was
not able to reliably differentiate between the pre and post-treatment
samples (advanced adenoma AUC range = 0.34 - 0.65). Interestingly, the
top 10 most important OTUs by MDA that were used for each model had
little overlap with each other {[}Figure 2{]}. Although treatment had an
impact on the overall community structure, the effect of treatment was
not consistent across patients and diagnosis groups. Both the adenoma
and carcioma treatment models had AUCs that were significantly higher
than a random model permutation (P-value \textless{} 0.0001).

@@

\textbf{\emph{Model Building.}} The Random Forest (33) algorithm was
used to create the three models used to classify pre and post-treatment
samples by diagnosis (adenoma, advanced adenoma, or carcinoma) as well
as to assess the probability that a sample was more similar to the
patient's original diagnosis or that of a disease-free patient. All
models included only OTU data obtained from 16S rRNA sequencing and were
processed using the caret (v6.0.76) R package. For each model we
optimized the mtry hyper-parameter, which defines the number of OTUs to
investigate at each split before a new division of the data was created
with the Random Forest model (33). To insure that our optimization did
not result in over-fitting of the data, we made 100 different 80/20
(train/test) splits of the data where the same proportion was present
within both the whole data set and the 80/20 split. For each of the 100
splits, 20 repeated 10-fold cross validation was performed on the 80\%
component to optimize the mtry hyper-parameter by maximizing the AUC
(Area Under the Curve of the Receiver Operator Characteristic). The
resulting model was then tested on the 20\% of the data that were held
out. A summary of the mtry hyperparameter values that were tried is
available in Table S5. The reported P-values for each model relative to
a random labeling was assessed by comparing the distribution of the 100
80/20 splits for the correctly labeled data to the distribution of
randomly labeled data.

The three diagnosis models were constructed by using the data from
Baxter et al. (24), which was censored for the pre-treatment samples of
the patients that we had post-treatment samples. The treatment models
were then used to quantify the model probability that a patient with an
initial diagnosis retained that diagnosis or a disease-free diagnosis.
